### 2023-10-26 Paper.Rmd
basic structure for document

---
title: "Urbanization and Public Transportation of Los Angeles"
author: "Kainoa Kanter, Julian Jacobson, Aidan Lastname"
date: "2023-10-25" # Change to date when we submit
output:
    pdf_document:
        extra_dependencies: ['amsmath']
---

inline equation $A = \pi r^2$

$$
f(x) = ax^2+bx+x
$$

### 2023-10-26 Paper.Rmd
Update Paper.Rmd

@@
 title: "Urbanization and Public Transportation of Los Angeles"
-author: "Kainoa Kanter, Julian Jacobson, Aidan Lastname"
-date: "2023-10-25" # Change to date when we submit
+bibliography: master.bib
+author: 
+  name: "Kainoa Kanter, Julian Jacobson, Aidan Lastname"
+  affiliation: "Math 28, Professor Smith, Santa Monica College"
+date: "`r format(Sys.time(), '%B %d, %Y')`"
+geometry: margin=1in
 output:

### 2023-10-26 Paper.Rmd
Abstract

--- 2023-10-26 Paper.Rmd
@@
 author: "Kainoa Kanter, Julian Jacobson, Aidan Lastname"
-date: "2023-10-25" # Change to date when we submit
+date: "`r format(Sys.time(), '%B %d, %Y')`"
+geometry: margin=1in
+abstract: "This paper will examine how the rapid growth of Los Angeles affects the demand for public transportation, utilizing derivatives to determine the rate of change in public transportation usage and studying the elasticity of demand for transportation options based on pricing, availability, convenience, and modes."
 output:

### 2023-11-06 data.r
add data

@@ -1,0 +1,11 @@
+library(readr)
+POPULATION_Areas_and_Densities_Table_1 <- read_csv("data/POPULATION_Areas_and_Densities_Table_1.csv")
+View(POPULATION_Areas_and_Densities_Table_1)
+city_revenue_by_month_transportation_20231106 <- read_csv("data/city_revenue_by_month_transportation_20231106.csv")
+View(city_revenue_by_month_transportation_20231106)
+CAFR_City_Activities_2018_Full_20231106 <- read_csv("data/CAFR_City_Activities_2018__Full__20231106.csv")
+View(CAFR_City_Activities_2018_Full_20231106)
+City_Budget_Expenditures_on_Transportation_20231106 <- read_csv("data/City_Budget_Expenditures_on_Transportation_20231106.csv")
+View(City_Budget_Expenditures_on_Transportation_20231106)
+Annual_Ridership <- read_csv("data/Annual Ridership.csv")
+View(Annual_Ridership)

### 2023-11-07 Paper.Rmd
Add my part

--- 2023-10-26 Paper.Rmd
@@
 title: "Urbanization and Public Transportation of Los Angeles"
-author: "Kainoa Kanter, Julian Jacobson, Aidan Lastname"
-date: "2023-10-25" # Change to date when we submit
+author: "Kainoa Kanter, Julian Jacobson, Aidan Allen"
+date: "`r format(Sys.time(), '%B %d, %Y')`"
+geometry: margin=1in
+abstract: "This paper will examine how the rapid growth of Los Angeles affects the demand for public transportation, utilizing derivatives to determine the rate of change in public transportation usage and studying the elasticity of demand for transportation options based on pricing, availability, convenience, and modes."
 output:
@@
 
-inline equation $A = \pi r^2$
+```{r setup, echo=FALSE, warning=FALSE, message=FALSE}
+knitr::opts_chunk$set(
+	echo = FALSE,
+	message = FALSE,
+	warning = FALSE
+)
+library(ggplot2)
+library(latex2exp)
+library(dplyr)
+library(tidyr)
+library(scales)
+library(stargazer)
+ridership_data <- read.csv("data/April_Transit_Ridership.csv")
+budget_data <- read.csv("data/City_Budget_Expenditures_on_Transportation_20231106.csv")
+```
+
+# Introduction
+
+Los Angeles -- the sprawling metropolis that we call home -- has become synonymous with both opportunity and urban challenges and has experienced rapid growth in recent decades.
+This growth has significant implications for various sectors, particularly the public transportation system.
+As one of the most car-dependent cities in the United States, Los Angeles faces unique challenges in scaling its public transportation network to meet increasing demand.
+Understanding the dynamics of public transportation usage and the factors influencing ridership is critical for planning and policy development.
+
+The aim of this study is to examine how the rapid growth of Los Angeles affects the demand for public transportation.
+The investigation includes a quantitative assessment utilizing derivatives to determine the rate of change in public transportation usage.
+Additionally, the study examines the elasticity of demand for transportation options in Los Angeles based on pricing, availability, convenience, and modes.
+This analysis offers insights into the responsiveness of ridership to changes in these factors, informing strategies to enhance the public transportation system.
+
+## Methodology
+
+### Data Collection
+
+The analysis is based on comprehensive datasets including average weekend and weekday ridership from 2019 to 2023, as well as a range of urban metrics such as crime rates, accident statistics, and city maintenance records.
+Yearly transportation budget figures also contribute to understanding the financial aspects influencing public transportation operations.
+
+The methodological approach consists of two primary quantitative analyses:
+
+#### Rate of Change Analysis:
+
+-   Utilize time-series data to calculate the first derivative of ridership numbers, yielding the rate of change over the years.
+This derivative analysis will be conducted separately for bus and rail services, offering a granular view of usage trends.
+-   Data transformation into a suitable format for analysis using the tidyverse collection of R packages.
+-   Visualization of trends through line plots, elucidating patterns and growth trajectories.
+
+#### Elasticity of Demand Analysis:
+
+-   Analyze the responsiveness of public transportation demand to changes in price (fare changes), availability (service hours, frequency), convenience (route changes, wait times), and modes (bus vs.
+rail).
+
+-   Employ econometric modeling to estimate the price elasticity of demand, cross-elasticity, and income elasticity, using regression analysis where ridership
+
+#### Visualization
+
+-   Data visualization was implemented using the `ggplot2` package in R
+-   Line plots were created to illustrate the trends in public transportation ridership over time, broken down by transportation type (bus and rail)
+-   Adjustments were made to display actual numbers instead of scientific notation for clarity
+
+\newpage
+
+# Exploratory Data Analysis
+
+```{r rate-of-change-plot, fig.cap="Line plot of rate of change in public transportation ridership over time"}
+ridership_long <- gather(ridership_data, key = "Type", value = "RateOfChange", -Year)
+
+# Create the line plot
+ggplot(ridership_long, aes(x=Year, y=RateOfChange, color=Type, group=Type)) +
+  geom_line() +
+  scale_y_continuous(labels = label_comma()) +
+  labs(title="Trend in Rate of Change of Public Transportation Usage",
+       x="Year",
+       y="Rate of Change in Total Monthly Ridership",
+       color="Transportation Type") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023))
+```
+
+\newpage
+
+# Derivative Analysis
+
+The rate of change in public transportation usage offers valuable insights into trends and can help forecast future demand.
+By differentiating the ridership data with respect to time, we can obtain the instantaneous rate of change, which is expressed mathematically as:
 
 $$
-f(x) = ax^2+bx+x
+R'(t) = \frac{dR}{dt}
 $$
+
+where $R(t)$ represents the ridership at time $t$. This derivative analysis provides the velocity of ridership change, indicating whether usage is increasing or decreasing over time.
+
+The following R code calculates the rate of change for Total Monthly Bus and Total Monthly Rail ridership and performs simple linear regression on these rates to understand how they change over the years:
+
+```{r rate-of-change}
+# Calculate the rate of change for Total Monthly Bus and Rail
+ridership_data$BusRateOfChange <- c(NA, diff(ridership_data$Total.Monthly.Bus))
+ridership_data$RailRateOfChange <- c(NA, diff(ridership_data$Total.Monthly.Rail))
+
+# Simple linear regression on the rate of change
+bus_model <- lm(BusRateOfChange ~ Year, data = ridership_data)
+rail_model <- lm(RailRateOfChange ~ Year, data = ridership_data)
+
+# Polynomial regression on the rate of change
+bus_model_poly <- lm(BusRateOfChange ~ poly(Year, 2, raw=TRUE), data = ridership_data)
+rail_model_poly <- lm(RailRateOfChange ~ poly(Year, 2, raw=TRUE), data = ridership_data)
+
+```
+
+```{r rate-of-change-results, results='asis'}
+stargazer(bus_model, rail_model, type = "latex", title = "Linear Regression Results", header=FALSE, omit=c("Year","Constant"))
+stargazer(bus_model_poly, rail_model_poly, type = "latex", title = "Polynomial Regression Results", header=FALSE, omit=c("poly","Constant"))
+```
+
+```{r predict-trend-usage}
+ridership_trend_long <- gather(ridership_data, key = "Type", value = "TotalRidership", -Year)
+
+# Predict the rate of change for the fitted models
+ridership_data$PredictedBusROC <- predict(bus_model, newdata=ridership_data)
+ridership_data$PredictedRailROC <- predict(rail_model, newdata=ridership_data)
+
+# Predict the rate of change for the fitted polynomial models
+ridership_data$PredictedBusROC_poly <- predict(bus_model_poly, newdata=ridership_data)
+ridership_data$PredictedRailROC_poly <- predict(rail_model_poly, newdata=ridership_data)
+
+```
+
+```{r trend-usage, fig.cap="Line plot of public transportation ridership over time with regression lines"}
+# Create the line plot with observed data
+p <- ggplot(ridership_trend_long, aes(x=Year, y=TotalRidership, color=Type, group=Type)) +
+  geom_line() +
+  # Assuming ridership_data has a 'Year' column and the predicted values only.
+  # Add the regression lines for predicted bus data (explicit aesthetics)
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC, group=1), color="blue", linetype="dashed", size=1) +
+  # Add the regression lines for predicted rail data (explicit aesthetics)
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC, group=1), color="red", linetype="dashed", size=1) +
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC_poly, group=2), color="hotpink", linetype="dotted", size=1) +
+  # Add the regression lines for predicted rail data (explicit aesthetics)
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC_poly, group=2), color="darkgreen", linetype="dotted", size=1) +
+  scale_y_continuous(labels = scales::label_comma()) +
+  labs(title="Trend in Public Transportation Usage",
+       x="Year",
+       y="Total Monthly Ridership",
+       color="Transportation Type") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023))
+
+# Data of 2023
+last_year <- max(ridership_data$Year)
+
+# Add labels for the dashed lines
+p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=-1) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC[ridership_data$Year == last_year], label="Predicted Rail Linear", color="red", hjust=1.1, vjust=4) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Bus Quadratic", color="hotpink", hjust=1.1, vjust=6) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkgreen", hjust=1.1, vjust=8)
+```

### 2023-11-07 Paper.Rmd
minor changes

@@
 geometry: margin=1in
-abstract: "This paper will examine how the rapid growth of Los Angeles affects the demand for public transportation, utilizing derivatives to determine the rate of change in public transportation usage and studying the elasticity of demand for transportation options based on pricing, availability, convenience, and modes."
+abstract: "This paper will examine how Los Angeles's demand for public transportation has been affected the last few years, utilizing derivatives to determine the rate of change in public transportation usage and studying the elasticity of demand for transportation options based on pricing, availability, convenience, and modes."
 output:
@@
 # Add labels for the dashed lines
-p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=-1) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC[ridership_data$Year == last_year], label="Predicted Rail Linear", color="red", hjust=1.1, vjust=4) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Bus Quadratic", color="hotpink", hjust=1.1, vjust=6) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkgreen", hjust=1.1, vjust=8)
+p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=6) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC[ridership_data$Year == last_year], label="Predicted Rail Linear", color="red", hjust=1.1, vjust=6) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Bus Quadratic", color="hotpink", hjust=1.1, vjust=8) +
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkgreen", hjust=1.1, vjust=10)
 ```

### 2023-11-07 data.r
Delete data.r

@@ -1,11 +1,0 @@
-library(readr)
-POPULATION_Areas_and_Densities_Table_1 <- read_csv("data/POPULATION_Areas_and_Densities_Table_1.csv")
-View(POPULATION_Areas_and_Densities_Table_1)
-city_revenue_by_month_transportation_20231106 <- read_csv("data/city_revenue_by_month_transportation_20231106.csv")
-View(city_revenue_by_month_transportation_20231106)
-CAFR_City_Activities_2018_Full_20231106 <- read_csv("data/CAFR_City_Activities_2018__Full__20231106.csv")
-View(CAFR_City_Activities_2018_Full_20231106)
-City_Budget_Expenditures_on_Transportation_20231106 <- read_csv("data/City_Budget_Expenditures_on_Transportation_20231106.csv")
-View(City_Budget_Expenditures_on_Transportation_20231106)
-Annual_Ridership <- read_csv("data/Annual Ridership.csv")
-View(Annual_Ridership)

### 2023-11-08 Paper.Rmd
graph styling

@@
   geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC, group=1), color="red", linetype="dashed", size=1) +
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC_poly, group=2), color="hotpink", linetype="dotted", size=1) +
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC_poly, group=2), color="hotpink", linetype="dotted", size=2) +
   # Add the regression lines for predicted rail data (explicit aesthetics)
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC_poly, group=2), color="darkgreen", linetype="dotted", size=1) +
-  scale_y_continuous(labels = scales::label_comma()) +
+  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC_poly, group=2), color="darkorange", linetype="dotted", size=2) +
+  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 16)) +
   labs(title="Trend in Public Transportation Usage",
@@
 # Add labels for the dashed lines
-p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=6) +
+p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=6.8) +
   annotate("text", x=last_year, y=ridership_data$PredictedRailROC[ridership_data$Year == last_year], label="Predicted Rail Linear", color="red", hjust=1.1, vjust=6) +
   annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Bus Quadratic", color="hotpink", hjust=1.1, vjust=8) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkgreen", hjust=1.1, vjust=10)
+  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkorange", hjust=1.1, vjust=10)
 ```

### 2023-11-08 Paper.Rmd
16 axis -> 14 axis

@@
   geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC_poly, group=2), color="darkorange", linetype="dotted", size=2) +
-  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 16)) +
+  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14)) +
   labs(title="Trend in Public Transportation Usage",

### 2023-11-09 Paper.Rmd
rm
Update Paper.Rmd

--- 2023-11-07 Paper.Rmd
@@
 ridership_data <- read.csv("data/April_Transit_Ridership.csv")
+population_data <- read.csv("data/Los_Angeles_Population.csv")
 budget_data <- read.csv("data/City_Budget_Expenditures_on_Transportation_20231106.csv")
@@
 Understanding the dynamics of public transportation usage and the factors influencing ridership is critical for planning and policy development.
+
+Testhfgg
 
@@
   labs(title="Trend in Rate of Change of Public Transportation Usage",
-       x="Year",
+       x="April of Each Year",
        y="Rate of Change in Total Monthly Ridership",
@@
 
-The following R code calculates the rate of change for Total Monthly Bus and Total Monthly Rail ridership and performs simple linear regression on these rates to understand how they change over the years:
-
-```{r rate-of-change}
+```{r rate-of-change-and-regression}
+# Derivative Analysis and Regression
 # Calculate the rate of change for Total Monthly Bus and Rail
@@
 
-```
-
-```{r rate-of-change-results, results='asis'}
-stargazer(bus_model, rail_model, type = "latex", title = "Linear Regression Results", header=FALSE, omit=c("Year","Constant"))
-stargazer(bus_model_poly, rail_model_poly, type = "latex", title = "Polynomial Regression Results", header=FALSE, omit=c("poly","Constant"))
-```
-
-```{r predict-trend-usage}
-ridership_trend_long <- gather(ridership_data, key = "Type", value = "TotalRidership", -Year)
-
 # Predict the rate of change for the fitted models
@@
 
+# Normalize the ridership data with the population data to get ridership per capita
+ridership_data <- ridership_data %>%
+  left_join(population_data, by = "Year") %>%
+  mutate(
+    RidershipPerCapitaBus = Total.Monthly.Bus / population,
+    RidershipPerCapitaRail = Total.Monthly.Rail / population
+  )
+
+# Calculate the rate of change for Ridership per Capita
+ridership_data$BusRateOfChangePerCapita <- c(NA, diff(ridership_data$RidershipPerCapitaBus))
+ridership_data$RailRateOfChangePerCapita <- c(NA, diff(ridership_data$RidershipPerCapitaRail))
+
+# Linear and polynomial regression for ridership per capita
+bus_model_per_capita <- lm(BusRateOfChangePerCapita ~ Year, data = ridership_data)
+rail_model_per_capita <- lm(RailRateOfChangePerCapita ~ Year, data = ridership_data)
 ```
 
-```{r trend-usage, fig.cap="Line plot of public transportation ridership over time with regression lines"}
-# Create the line plot with observed data
-p <- ggplot(ridership_trend_long, aes(x=Year, y=TotalRidership, color=Type, group=Type)) +
-  geom_line() +
-  # Assuming ridership_data has a 'Year' column and the predicted values only.
-  # Add the regression lines for predicted bus data (explicit aesthetics)
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC, group=1), color="blue", linetype="dashed", size=1) +
-  # Add the regression lines for predicted rail data (explicit aesthetics)
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC, group=1), color="red", linetype="dashed", size=1) +
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedBusROC_poly, group=2), color="hotpink", linetype="dotted", size=1) +
-  # Add the regression lines for predicted rail data (explicit aesthetics)
-  geom_line(data=ridership_data, aes(x=Year, y=PredictedRailROC_poly, group=2), color="darkgreen", linetype="dotted", size=1) +
+```{r regression-results, results='asis'}
+# Regression Results
+stargazer(bus_model, rail_model, type = "latex", title = "Linear Regression Results", header=FALSE, omit=c("Year","Constant"))
+stargazer(bus_model_poly, rail_model_poly, type = "latex", title = "Polynomial Regression Results", header=FALSE, omit=c("poly","Constant"))
+```
+
+```{r predict-trend-usage}
+# Predict the rate of change per capita for all years using the linear and polynomial models
+# (This assumes you have a model fitted to per capita data)
+all_years <- data.frame(Year = seq(min(ridership_data$Year), 2027))
+all_years <- left_join(all_years, population_data, by = "Year")
+
+all_years$PredictedBusROCPerCapita <- predict(bus_model_per_capita, newdata=all_years)
+all_years$PredictedRailROCPerCapita <- predict(rail_model_per_capita, newdata=all_years)
+
+# Assuming the rate of change per capita is per year, integrate the rate of change to get the total ridership per capita
+first_known_bus_ridership_per_capita <- head(ridership_data$RidershipPerCapitaBus, 1)
+first_known_rail_ridership_per_capita <- head(ridership_data$RidershipPerCapitaRail, 1)
+
+all_years$RidershipPerCapitaBus <- first_known_bus_ridership_per_capita + cumsum(c(0, all_years$PredictedBusROCPerCapita[-1]))
+all_years$RidershipPerCapitaRail <- first_known_rail_ridership_per_capita + cumsum(c(0, all_years$PredictedRailROCPerCapita[-1]))
+```
+
+```{r format-and-combine-data}
+# Format and Combine Data
+# Convert the predicted data to long format
+predicted_data_long <- all_years %>%
+  select(Year, RidershipPerCapitaBus, RidershipPerCapitaRail) %>%
+  pivot_longer(cols = c("RidershipPerCapitaBus", "RidershipPerCapitaRail"),
+               names_to = "Type",
+               values_to = "TotalRidership") %>%
+  mutate(Prediction = "Predicted")
+
+# Now, make sure the actual data is in the same long format with the same column names
+actual_data_long <- ridership_data %>%
+  select(Year, RidershipPerCapitaBus, RidershipPerCapitaRail) %>%
+  pivot_longer(cols = c("RidershipPerCapitaBus", "RidershipPerCapitaRail"),
+               names_to = "Type",
+               values_to = "TotalRidership") %>%
+  mutate(Prediction = "Actual")
+
+# Combine the actual and predicted data
+combined_data_long <- rbind(actual_data_long, predicted_data_long)
+
+# Filter out the extreme predicted values for the polynomial bus data
+combined_data_long <- combined_data_long %>%
+  filter(!(Type == "RidershipPerCapitaBus" & TotalRidership < -10000000 & Prediction == "Predicted"))
+
+```
+
+```{r plot-trend-usage, fig.cap="Line plot of public transportation ridership over time with regression lines"}
+# Create the line plot with observed and predicted per capita data
+p <- ggplot(combined_data_long, aes(x = Year, y = RidershipPerCapita, color = Type)) +
+  geom_line(aes(linetype = Prediction)) +
   scale_y_continuous(labels = scales::label_comma()) +
-  labs(title="Trend in Public Transportation Usage",
-       x="Year",
-       y="Total Monthly Ridership",
-       color="Transportation Type") +
-  scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023))
+  labs(title = "Trend in Public Transportation Usage Per Capita",
+       x = "April of Each Year",
+       y = "Ridership Per Capita",
+       color = "Transportation Type",
+       linetype = "Data Type") +
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(min(ridership_data$Year), max(all_years$Year)))
 
-# Data of 2023
-last_year <- max(ridership_data$Year)
+# Print the plot
+print(p)
 
-# Add labels for the dashed lines
-p + annotate("text", x=last_year, y=ridership_data$PredictedBusROC[ridership_data$Year == last_year], label="Predicted Bus Linear", color="blue", hjust=1.1, vjust=6) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC[ridership_data$Year == last_year], label="Predicted Rail Linear", color="red", hjust=1.1, vjust=6) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Bus Quadratic", color="hotpink", hjust=1.1, vjust=8) +
-  annotate("text", x=last_year, y=ridership_data$PredictedRailROC_poly[ridership_data$Year == last_year], label="Predicted Rail Quadratic", color="darkgreen", hjust=1.1, vjust=10)
 ```
+
+\newpage
+
+# Sources:
+
+- Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler
+
+- Los Angeles Population Data (real and predicted): WorldPopulationReview "Los Angeles, California Population"

### 2023-11-09 Paper.Rmd
revert julian test
Revert "Initial population data"
This reverts commit 6d8f0782ef3a20d55b434ed1a540f95499aec9f5.

@@
 ridership_data <- read.csv("data/April_Transit_Ridership.csv")
-population_data <- read.csv("data/Los_Angeles_Population.csv")
 budget_data <- read.csv("data/City_Budget_Expenditures_on_Transportation_20231106.csv")
@@
 Understanding the dynamics of public transportation usage and the factors influencing ridership is critical for planning and policy development.
-
-Testhfgg
 
@@
 ridership_data$PredictedRailROC_poly <- predict(rail_model_poly, newdata=ridership_data)
-
-# Normalize the ridership data with the population data to get ridership per capita
-ridership_data <- ridership_data %>%
-  left_join(population_data, by = "Year") %>%
-  mutate(
-    RidershipPerCapitaBus = Total.Monthly.Bus / population,
-    RidershipPerCapitaRail = Total.Monthly.Rail / population
-  )
-
-# Calculate the rate of change for Ridership per Capita
-ridership_data$BusRateOfChangePerCapita <- c(NA, diff(ridership_data$RidershipPerCapitaBus))
-ridership_data$RailRateOfChangePerCapita <- c(NA, diff(ridership_data$RidershipPerCapitaRail))
-
-# Linear and polynomial regression for ridership per capita
-bus_model_per_capita <- lm(BusRateOfChangePerCapita ~ Year, data = ridership_data)
-rail_model_per_capita <- lm(RailRateOfChangePerCapita ~ Year, data = ridership_data)
 ```
@@
 ```{r predict-trend-usage}
-# Predict the rate of change per capita for all years using the linear and polynomial models
-# (This assumes you have a model fitted to per capita data)
+# Predict Trend Usage
+# Extend the data frame to include all years for prediction
 all_years <- data.frame(Year = seq(min(ridership_data$Year), 2027))
-all_years <- left_join(all_years, population_data, by = "Year")
 
-all_years$PredictedBusROCPerCapita <- predict(bus_model_per_capita, newdata=all_years)
-all_years$PredictedRailROCPerCapita <- predict(rail_model_per_capita, newdata=all_years)
+# Predict the rate of change for all years using the linear and polynomial models
+all_years$PredictedBusROC <- predict(bus_model, newdata=all_years)
+all_years$PredictedRailROC <- predict(rail_model, newdata=all_years)
+all_years$PredictedBusROC_poly <- predict(bus_model_poly, newdata=all_years)
+all_years$PredictedRailROC_poly <- predict(rail_model_poly, newdata=all_years)
 
-# Assuming the rate of change per capita is per year, integrate the rate of change to get the total ridership per capita
-first_known_bus_ridership_per_capita <- head(ridership_data$RidershipPerCapitaBus, 1)
-first_known_rail_ridership_per_capita <- head(ridership_data$RidershipPerCapitaRail, 1)
+# Initialize the first known ridership value
+first_known_bus_ridership <- head(ridership_data$Total.Monthly.Bus, 1)
+first_known_rail_ridership <- head(ridership_data$Total.Monthly.Rail, 1)
 
-all_years$RidershipPerCapitaBus <- first_known_bus_ridership_per_capita + cumsum(c(0, all_years$PredictedBusROCPerCapita[-1]))
-all_years$RidershipPerCapitaRail <- first_known_rail_ridership_per_capita + cumsum(c(0, all_years$PredictedRailROCPerCapita[-1]))
+# Assuming the rate of change is per year, integrate the rate of change to get the total ridership
+# We start from the first known ridership value and accumulate the predicted rate of change
+all_years$Total.Monthly.Bus <- first_known_bus_ridership + cumsum(c(0, all_years$PredictedBusROC[-1]))
+all_years$Total.Monthly.Rail <- first_known_rail_ridership + cumsum(c(0, all_years$PredictedRailROC[-1]))
+all_years$Total.Monthly.Bus_poly <- first_known_bus_ridership + cumsum(c(0, all_years$PredictedBusROC_poly[-1]))
+all_years$Total.Monthly.Rail_poly <- first_known_rail_ridership + cumsum(c(0, all_years$PredictedRailROC_poly[-1]))
 ```
@@
 predicted_data_long <- all_years %>%
-  select(Year, RidershipPerCapitaBus, RidershipPerCapitaRail) %>%
-  pivot_longer(cols = c("RidershipPerCapitaBus", "RidershipPerCapitaRail"),
+  select(Year, Total.Monthly.Bus, Total.Monthly.Rail, Total.Monthly.Bus_poly, Total.Monthly.Rail_poly) %>%
+  pivot_longer(cols = c("Total.Monthly.Bus", "Total.Monthly.Rail", "Total.Monthly.Bus_poly", "Total.Monthly.Rail_poly"),
                names_to = "Type",
@@
 actual_data_long <- ridership_data %>%
-  select(Year, RidershipPerCapitaBus, RidershipPerCapitaRail) %>%
-  pivot_longer(cols = c("RidershipPerCapitaBus", "RidershipPerCapitaRail"),
+  select(Year, Total.Monthly.Bus, Total.Monthly.Rail) %>%
+  pivot_longer(cols = c("Total.Monthly.Bus", "Total.Monthly.Rail"),
                names_to = "Type",
@@
 combined_data_long <- combined_data_long %>%
-  filter(!(Type == "RidershipPerCapitaBus" & TotalRidership < -10000000 & Prediction == "Predicted"))
+  filter(!(Type == "Total.Monthly.Bus_poly" & TotalRidership < -10000000 & Prediction == "Predicted"))
 
@@
 ```{r plot-trend-usage, fig.cap="Line plot of public transportation ridership over time with regression lines"}
-# Create the line plot with observed and predicted per capita data
-p <- ggplot(combined_data_long, aes(x = Year, y = RidershipPerCapita, color = Type)) +
+# Plot Trend Usage
+# Create the line plot with observed data
+p <- ggplot(combined_data_long, aes(x = Year, y = TotalRidership, color = Type)) +
   geom_line(aes(linetype = Prediction)) +
   scale_y_continuous(labels = scales::label_comma()) +
-  labs(title = "Trend in Public Transportation Usage Per Capita",
+  labs(title = "Trend in Public Transportation Usage",
        x = "April of Each Year",
-       y = "Ridership Per Capita",
+       y = "Total Monthly Ridership",
        color = "Transportation Type",
        linetype = "Data Type") +
-  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(min(ridership_data$Year), max(all_years$Year)))
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(min(ridership_data$Year), max(all_years$Year))) +
+  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14))
 
-# Print the plot
+# Add predicted lines for past years (where we have actual data)
+predicted_data_long <- all_years %>%
+  select(Year, Total.Monthly.Bus, Total.Monthly.Rail, Total.Monthly.Bus_poly, Total.Monthly.Rail_poly) %>%
+  pivot_longer(cols = c("Total.Monthly.Bus", "Total.Monthly.Rail", "Total.Monthly.Bus_poly", "Total.Monthly.Rail_poly"),
+               names_to = "Type",
+               values_to = "TotalRidership") %>%
+  mutate(Prediction = "Predicted") %>%
+  filter(Year <= max(ridership_data$Year))
+
+# Overlay this on the plot with a different line type
+p <- p +
+  geom_line(data = predicted_data_long, aes(x = Year, y = TotalRidership, color = Type, linetype = Prediction))
+
 print(p)
-
 ```
-
-\newpage
-
-# Sources:
-
-- Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler
-
-- Los Angeles Population Data (real and predicted): WorldPopulationReview "Los Angeles, California Population"

### 2023-11-09 Paper.Rmd
Add mini tangent lines

@@
 # Create the line plot
-ggplot(ridership_long, aes(x=Year, y=RateOfChange, color=Type, group=Type)) +
+p <- ggplot(ridership_long, aes(x=Year, y=RateOfChange, color=Type, group=Type)) +
   geom_line() +
@@
   scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023))
+
+# Theme
+p <- p + theme_bw()
+
+# Show graph
+print(p)
 ```
@@
 
+y_feasible_min <- -10000000
+y_feasible_max <- 70000000
+
 # Filter out the extreme predicted values for the polynomial bus data
 combined_data_long <- combined_data_long %>%
-  filter(!(Type == "Total.Monthly.Bus_poly" & TotalRidership < -10000000 & Prediction == "Predicted"))
+  filter(!(Type == "Total.Monthly.Bus_poly" & TotalRidership < y_feasible_min & Prediction == "Predicted")) %>%
+  filter(!(Type == "Total.Monthly.Bus" & TotalRidership > y_feasible_max & Prediction == "Predicted"))
 
+```
+
+```{r add-tangent-lines}
+# Function to calculate tangent lines for a linear model
+add_tangent_lines <- function(plot, data, x_var, y_var, color, line_length = 0.5) {
+  # Fit a linear model
+  model <- lm(reformulate(x_var, y_var), data = data)
+
+  # Calculate the slopes (coefficients)
+  coef <- model$coefficients
+
+  # Create a new data frame for tangent lines
+  tangent_data <- data.frame(
+    x = data[[x_var]],
+    y = data[[y_var]],
+    slope = coef[2]
+  )
+
+  # Calculate y intercept for each tangent line
+  tangent_data$intercept <- with(tangent_data, y - slope * x)
+
+  # Create the start and end points for each tangent line
+  tangent_data$xstart <- tangent_data$x - line_length / 2
+  tangent_data$xend <- tangent_data$x + line_length / 2
+  tangent_data$ystart <- with(tangent_data, intercept + slope * xstart)
+  tangent_data$yend <- with(tangent_data, intercept + slope * xend)
+
+  tangent_data <- tangent_data %>%
+    filter(yend < y_feasible_max & yend > -y_feasible_min)
+
+  # Add the tangent lines to the plot
+  plot + geom_segment(data = tangent_data, aes(x = xstart, y = ystart, xend = xend, yend = yend), color = color, linetype = "solid", alpha=0.4)
+}
 ```
@@
 
+# Overlay real data
+p <- p +
+  geom_line(data = ridership_long, aes(x = Year, y = RateOfChange, color = Type), size=1)
+
 # Add predicted lines for past years (where we have actual data)
@@
 
+# Add tangent lines
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Bus", "red")
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Bus_poly", "darkgreen", line_length = 0.2)
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail", "turquoise", line_length=0.7)
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail_poly", "purple")
+
+# Theme
+p <- p + theme_bw()
+
+# Show graph
 print(p)
 ```
+
+\newpage
+
+# Sources:
+
+- Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler

### 2023-11-10 Paper.Rmd
cleanup

@@
 p <- ggplot(combined_data_long, aes(x = Year, y = TotalRidership, color = Type)) +
-  geom_line(aes(linetype = Prediction)) +
+  geom_line(data = filter(combined_data_long, Prediction == "Actual"), aes(linetype = Prediction)) +
+  geom_smooth(data = filter(combined_data_long, Prediction == "Predicted"), aes(linetype = Prediction), se=FALSE) +
   scale_y_continuous(labels = scales::label_comma()) +
@@
   scale_x_continuous(breaks = scales::pretty_breaks(n = 10), limits = c(min(ridership_data$Year), max(all_years$Year))) +
-  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14))
+  scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14)) +
+  scale_color_discrete(labels=c("Bus - Linear Reg.", "Bus - Quadratic Reg.", "Rail - Linear Reg.", "Bus - Quadratic Reg."))
 
@@
 
-# Add predicted lines for past years (where we have actual data)
-predicted_data_long <- all_years %>%
-  select(Year, Total.Monthly.Bus, Total.Monthly.Rail, Total.Monthly.Bus_poly, Total.Monthly.Rail_poly) %>%
-  pivot_longer(cols = c("Total.Monthly.Bus", "Total.Monthly.Rail", "Total.Monthly.Bus_poly", "Total.Monthly.Rail_poly"),
-               names_to = "Type",
-               values_to = "TotalRidership") %>%
-  mutate(Prediction = "Predicted") %>%
-  filter(Year <= max(ridership_data$Year))
-
-# Overlay this on the plot with a different line type
-p <- p +
-  geom_line(data = predicted_data_long, aes(x = Year, y = TotalRidership, color = Type, linetype = Prediction))
-
 # Add tangent lines
@@
 p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Bus_poly", "darkgreen", line_length = 0.2)
-p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail", "turquoise", line_length=0.7)
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail", "darkblue", line_length=0.7)
 p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail_poly", "purple")

### 2023-11-10 Paper.Rmd
Print equations, update gitignore

@@
 
+```{r format-equations}
+# Get the coefficients for the linear models
+linear_bus_coefs <- coef(bus_model)
+linear_rail_coefs <- coef(rail_model)
+
+# Get the coefficients for the polynomial models
+poly_bus_coefs <- coef(bus_model_poly)
+poly_rail_coefs <- coef(rail_model_poly)
+
+# Construct the linear equations as strings
+linear_bus_eq <- paste0("Total.Monthly.Bus = ", round(linear_bus_coefs[1], 2), " + (", round(linear_bus_coefs[2], 2), " * Year)")
+linear_rail_eq <- paste0("Total.Monthly.Rail = ", round(linear_rail_coefs[1], 2), " + (", round(linear_rail_coefs[2], 2), " * Year)")
+
+# Construct the quadratic equations as strings
+# Assuming the polynomial regression is a second-degree polynomial
+poly_bus_eq <- paste0("Total.Monthly.Bus = ", round(poly_bus_coefs[1], 2), " + (", round(poly_bus_coefs[2], 2), " * Year) + (", round(poly_bus_coefs[3], 2), " * Year^2)")
+poly_rail_eq <- paste0("Total.Monthly.Rail = ", round(poly_rail_coefs[1], 2), " + (", round(poly_rail_coefs[2], 2), " * Year) + (", round(poly_rail_coefs[3], 2), " * Year^2)")
+
+linear_bus_adjusted_intercept <- linear_bus_coefs[1] + linear_bus_coefs[2] * 2019
+linear_rail_adjusted_intercept <- linear_rail_coefs[1] + linear_rail_coefs[2] * 2019
+
+# For polynomial models
+# Note: This requires a bit more calculation because the year term is squared
+poly_bus_adjusted_intercept <- poly_bus_coefs[1] + poly_bus_coefs[2] * 2019 + poly_bus_coefs[3] * 2019^2
+poly_rail_adjusted_intercept <- poly_rail_coefs[1] + poly_rail_coefs[2] * 2019 + poly_rail_coefs[3] * 2019^2
+```
+
+```{r show-equations, results='asis'}
+# Define the equations as LaTeX strings
+linear_bus_eq_adjusted <- sprintf("\\text{Total\\ Monthly\\ Bus} = %.2f + %.2f \\cdot (Year - 2019)", 
+                                  linear_bus_adjusted_intercept, linear_bus_coefs[2])
+linear_rail_eq_adjusted <- sprintf("\\text{Total\\ Monthly\\ Rail} = %.2f + %.2f \\cdot (Year - 2019)", 
+                                   linear_rail_adjusted_intercept, linear_rail_coefs[2])
+
+poly_bus_eq_adjusted <- sprintf("\\text{Total\\ Monthly\\ Bus} = %.2f + %.2f \\cdot (Year - 2019) + %.2f \\cdot (Year - 2019)^2", 
+                                poly_bus_adjusted_intercept, poly_bus_coefs[2], poly_bus_coefs[3])
+poly_rail_eq_adjusted <- sprintf("\\text{Total\\ Monthly\\ Rail} = %.2f + %.2f \\cdot (Year - 2019) + %.2f \\cdot (Year - 2019)^2", 
+                                 poly_rail_adjusted_intercept, poly_rail_coefs[2], poly_rail_coefs[3])
+
+cat("$$", linear_bus_eq_adjusted, "$$")
+cat("$$", linear_rail_eq_adjusted, "$$")
+cat("$$", poly_bus_eq_adjusted, "$$")
+cat("$$", poly_rail_eq_adjusted, "$$")
+```
+
 ```{r format-and-combine-data}
@@
   geom_line(data = filter(combined_data_long, Prediction == "Actual"), aes(linetype = Prediction)) +
-  geom_smooth(data = filter(combined_data_long, Prediction == "Predicted"), aes(linetype = Prediction), se=FALSE) +
+  geom_smooth(data = filter(combined_data_long, Prediction == "Predicted"), aes(linetype = Prediction), size=0.6, se=FALSE) +
   scale_y_continuous(labels = scales::label_comma()) +

### 2023-11-10 Paper.Rmd
revert darkblue back to turquoise

@@
 p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Bus_poly", "darkgreen", line_length = 0.2)
-p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail", "darkblue", line_length=0.7)
+p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail", "turquoise", line_length=0.7)
 p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail_poly", "purple")

### 2023-11-10 Paper.Rmd
label bus and rail

@@
        color="Transportation Type") +
-  scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023))
+  scale_x_continuous(breaks = scales::pretty_breaks(n = 5), limits = c(2019, 2023)) +
+  scale_color_discrete(labels=c("Bus", "Rail"))
 

### 2023-11-22 Paper.tex
prep

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{amsmath}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Urbanization and Public Transportation of Los Angeles},
  pdfauthor={Kainoa Kanter, Julian Jacobson, Aidan Allen},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Urbanization and Public Transportation of Los Angeles}
\author{Kainoa Kanter, Julian Jacobson, Aidan Allen}
\date{November 21, 2023}

\begin{document}
\maketitle
\begin{abstract}
This paper will examine how Los Angeles's demand for public
transportation has been affected the last few years, utilizing
derivatives to determine the rate of change in public transportation
usage and studying the elasticity of demand for transportation options
based on pricing, availability, convenience, and modes.
\end{abstract}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Los Angeles -- the sprawling metropolis that we call home -- has become
synonymous with both opportunity and urban challenges and has
experienced rapid growth in recent decades. This growth has significant
implications for various sectors, particularly the public transportation
system. As one of the most car-dependent cities in the United States,
Los Angeles faces unique challenges in scaling its public transportation
network to meet increasing demand. Understanding the dynamics of public
transportation usage and the factors influencing ridership is critical
for planning and policy development.

The aim of this study is to examine how the rapid growth of Los Angeles
affects the demand for public transportation. The investigation includes
a quantitative assessment utilizing derivatives to determine the rate of
change in public transportation usage. Additionally, the study examines
the elasticity of demand for transportation options in Los Angeles based
on pricing, availability, convenience, and modes. This analysis offers
insights into the responsiveness of ridership to changes in these
factors, informing strategies to enhance the public transportation
system.

\hypertarget{methodology}{%
\subsection{Methodology}\label{methodology}}

\hypertarget{data-collection}{%
\subsubsection{Data Collection}\label{data-collection}}

The analysis is based on comprehensive datasets including average
weekend and weekday ridership from 2019 to 2023, as well as a range of
urban metrics such as crime rates, accident statistics, and city
maintenance records. Yearly transportation budget figures also
contribute to understanding the financial aspects influencing public
transportation operations.

The methodological approach consists of two primary quantitative
analyses:

\hypertarget{rate-of-change-analysis}{%
\paragraph{Rate of Change Analysis:}\label{rate-of-change-analysis}}

\begin{itemize}
\tightlist
\item
  Utilize time-series data to calculate the first derivative of
  ridership numbers, yielding the rate of change over the years. This
  derivative analysis will be conducted separately for bus and rail
  services, offering a granular view of usage trends.
\item
  Data transformation into a suitable format for analysis using the
  tidyverse collection of R packages.
\item
  Visualization of trends through line plots, elucidating patterns and
  growth trajectories.
\end{itemize}

\hypertarget{elasticity-of-demand-analysis}{%
\paragraph{Elasticity of Demand
Analysis:}\label{elasticity-of-demand-analysis}}

\begin{itemize}
\item
  Analyze the responsiveness of public transportation demand to changes
  in price (fare changes), availability (service hours, frequency),
  convenience (route changes, wait times), and modes (bus vs.~rail).
\item
  Employ econometric modeling to estimate the price elasticity of
  demand, cross-elasticity, and income elasticity, using regression
  analysis where ridership
\end{itemize}

\hypertarget{visualization}{%
\paragraph{Visualization}\label{visualization}}

\begin{itemize}
\tightlist
\item
  Data visualization was implemented using the \texttt{ggplot2} package
  in R
\item
  Line plots were created to illustrate the trends in public
  transportation ridership over time, broken down by transportation type
  (bus and rail)
\item
  Adjustments were made to display actual numbers instead of scientific
  notation for clarity
\end{itemize}

\newpage

\hypertarget{exploratory-data-analysis}{%
\section{Exploratory Data Analysis}\label{exploratory-data-analysis}}

\begin{figure}
\centering
\includegraphics{Paper_files/figure-latex/rate-of-change-plot-1.pdf}
\caption{Line plot of rate of change in public transportation ridership
over time}
\end{figure}

\newpage

\hypertarget{derivative-analysis}{%
\section{Derivative Analysis}\label{derivative-analysis}}

The rate of change in public transportation usage offers valuable
insights into trends and can help forecast future demand. By
differentiating the ridership data with respect to time, we can obtain
the instantaneous rate of change, which is expressed mathematically as:

\[
R'(t) = \frac{dR}{dt}
\]

where \(R(t)\) represents the ridership at time \(t\). This derivative
analysis provides the velocity of ridership change, indicating whether
usage is increasing or decreasing over time.

\begin{table}[!htbp] \centering 
  \caption{Linear Regression Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{2}{c}{\textit{Dependent variable:}} \\ 
\cline{2-3} 
\\[-1.8ex] & BusRateOfChange & RailRateOfChange \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
\hline \\[-1.8ex] 
Observations & 4 & 4 \\ 
R$^{2}$ & 0.319 & 0.184 \\ 
Adjusted R$^{2}$ & $-$0.021 & $-$0.223 \\ 
Residual Std. Error (df = 2) & 13,934,596.000 & 2,498,636.000 \\ 
F Statistic (df = 1; 2) & 0.938 & 0.452 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\begin{table}[!htbp] \centering 
  \caption{Polynomial Regression Results} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{2}{c}{\textit{Dependent variable:}} \\ 
\cline{2-3} 
\\[-1.8ex] & BusRateOfChange & RailRateOfChange \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
\hline \\[-1.8ex] 
Observations & 4 & 4 \\ 
R$^{2}$ & 0.806 & 0.243 \\ 
Adjusted R$^{2}$ & 0.418 & $-$1.272 \\ 
Residual Std. Error (df = 1) & 10,522,144.000 & 3,405,367.000 \\ 
F Statistic (df = 2; 1) & 2.076 & 0.160 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table}

\[ \text{Total\ Monthly\ Bus} = -16456298.50 + 6036028.40 \cdot (Year - 2019) \]\[ \text{Total\ Monthly\ Rail} = -2597795.50 + 751531.80 \cdot (Year - 2019) \]\[ \text{Total\ Monthly\ Bus} = -58111913.49 + 33688766315.52 \cdot (Year - 2019) + -8331123.00 \cdot (Year - 2019)^2 \]\[ \text{Total\ Monthly\ Rail} = -239513.00 + -1906155697.79 \cdot (Year - 2019) + 471656.50 \cdot (Year - 2019)^2 \]

\begin{figure}
\centering
\includegraphics{Paper_files/figure-latex/plot-trend-usage-1.pdf}
\caption{Line plot of public transportation ridership over time with
regression lines}
\end{figure}

\newpage

\hypertarget{sources}{%
\section{Sources:}\label{sources}}

\begin{itemize}
\tightlist
\item
  Transit ridership data: Los Angeles Metro ``L.A. METRO TRANSIT
  RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD'', Patrick Chandler
\end{itemize}

\end{document}

### 2023-11-22 Paper.Rmd
prep
Update Paper.Rmd

--- 2023-11-10 Paper.Rmd
@@
 where $R(t)$ represents the ridership at time $t$. This derivative analysis provides the velocity of ridership change, indicating whether usage is increasing or decreasing over time.
+
+It is important to keep in mind the usage of information like such.
+For Metro, getting their ridership to pre-pandemic numbers would be a huge accomplishment in the public transit landscape of Los Angeles.
+Total ridership in May 2023 was at 77% of 2019 levels, and in June 2023, it was at 81% of its 2019 pre-pandemic level.
+When using March of 2020 as our reference point, we can analyze the potentially differing rates of change in the ridership of Metro Bus and Metro Rail usage over time.
+Over the first year from 2020, the total monthly ridership went down roughly by roughly 22.5 million users.
+
+\
+
+The derivative of bus rider usage in Los Angeles from 2019 to 2022 shows a gradual recovery in ridership, with consistent year-over-year increases and efforts to restore pre-pandemic levels.
+The recovery of bus ridership in Los Angeles reflects the ongoing efforts of Metro to improve mobility and public transit accessibility in the region.
+These changes indicate a positive trend in bus rider usage, with a focus on restoring and surpassing pre-pandemic ridership levels.
 

### 2023-11-22 Paper.Rmd
Fix label

@@
   scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14)) +
-  scale_color_discrete(labels=c("Bus - Linear Reg.", "Bus - Quadratic Reg.", "Rail - Linear Reg.", "Bus - Quadratic Reg."))
+  scale_color_discrete(labels=c("Bus - Linear Reg.", "Bus - Quadratic Reg.", "Rail - Linear Reg.", "Rail - Quadratic Reg."))
 

### 2023-11-22 Paper.Rmd
Add unused gganimate code

--- 2023-11-22 Paper.Rmd
@@
 date: "`r format(Sys.time(), '%B %d, %Y')`"
-geometry: margin=1in
-abstract: "This paper will examine how Los Angeles's demand for public transportation has been affected the last few years, utilizing derivatives to determine the rate of change in public transportation usage and studying the elasticity of demand for transportation options based on pricing, availability, convenience, and modes."
 output:
   pdf_document:
-    extra_dependencies: ['amsmath']
+    extra_dependencies: amsmath
+  html_document:
+    df_print: paged
+abstract: This paper will examine how Los Angeles's demand for public transportation
+  has been affected the last few years, utilizing derivatives to determine the rate
+  of change in public transportation usage and studying the elasticity of demand for
+  transportation options based on pricing, availability, convenience, and modes.
+geometry: margin=1in
 ---
@@
 library(stargazer)
+#library(gganimate)
 ridership_data <- read.csv("data/April_Transit_Ridership.csv")
@@
   scale_y_continuous(labels = scales::label_comma(), breaks = scales::pretty_breaks(n = 14)) +
-  scale_color_discrete(labels=c("Bus - Linear Reg.", "Bus - Quadratic Reg.", "Rail - Linear Reg.", "Bus - Quadratic Reg."))
+  scale_color_discrete(labels=c("Bus - Linear Reg.", "Bus - Quadratic Reg.", "Rail - Linear Reg.", "Rail - Quadratic Reg.")) +
+  #transition_states(Type, transition_length = 2, state_length = 1) +
+  theme_bw()
 
@@
   geom_line(data = ridership_long, aes(x = Year, y = RateOfChange, color = Type), size=1)
+
+#animation <- animate(
+#  p,
+#  fps = 10,
+#  duration = 10,
+#  width = 800,
+#  height = 600
+#)
+#anim_save("animation.gif")
+#unlink("frames", recursive = TRUE)
 
@@
 p <- add_tangent_lines(p, all_years, "Year", "Total.Monthly.Rail_poly", "purple")
-
-# Theme
-p <- p + theme_bw()
 

### 2023-11-27 Paper.Rmd
Update Paper.Rmd

@@
 This analysis offers insights into the responsiveness of ridership to changes in these factors, informing strategies to enhance the public transportation system.
+
+This information is crucial for making impactful decisions related to resource allocation, service expansion, and policy adjustments.
+Transit agencies can use this information to optimize fare structures, implement dynamic pricing models, or introduce discounts and incentives to attract more riders during specific times or for particular modes.
+Additionally, knowing how pricing, availability, and convenience impact ridership allows for targeted marketing and communication strategies.
 

### 2023-11-27 Paper.Rmd
knit

@@
 output:
+  html_document:
+    df_print: paged
   pdf_document:
     extra_dependencies: amsmath
-  html_document:
-    df_print: paged
 abstract: This paper will examine how Los Angeles's demand for public transportation

### 2023-11-27 Paper.Rmd
Better exploratory analysis

@@
   has been affected the last few years, utilizing derivatives to determine the rate
-  of change in public transportation usage and studying the elasticity of demand for
-  transportation options based on pricing, availability, convenience, and modes.
+  of change in public transportation usage.
 geometry: margin=1in
@@
 
-#### Elasticity of Demand Analysis:
-
--   Analyze the responsiveness of public transportation demand to changes in price (fare changes), availability (service hours, frequency), convenience (route changes, wait times), and modes (bus vs.
-rail).
-
--   Employ econometric modeling to estimate the price elasticity of demand, cross-elasticity, and income elasticity, using regression analysis where ridership
-
 #### Visualization
@@
 
-\newpage
-
 # Exploratory Data Analysis
 
-```{r rate-of-change-plot, fig.cap="Line plot of rate of change in public transportation ridership over time"}
+As seen in the graph below, ridership data for both the Metro Bus system and Metro Light Rail between 2019 and 2023 has been heavily impacted by the 2020 pandemic.
+The bus ridership has had more of a sharp decline from 2019 to 2020 and bounce-back shortly, while the rail was slowing down from 2019 - 2021, with not a lot of recovery.
+
+```{r rate-of-change-plot, fig.cap="Line plot of rate of change in public transportation ridership over time", fig.height=4}
 ridership_long <- gather(ridership_data, key = "Type", value = "RateOfChange", -Year)
@@
 -   Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler
+-   R Paper source code: https://github.com/thatonecalculator/math-28-project

### 2023-11-29 Paper.Rmd
don't mention eod for now

@@
 The investigation includes a quantitative assessment utilizing derivatives to determine the rate of change in public transportation usage.
-Additionally, the study examines the elasticity of demand for transportation options in Los Angeles based on pricing, availability, convenience, and modes.
 This analysis offers insights into the responsiveness of ridership to changes in these factors, informing strategies to enhance the public transportation system.
 
-This information is crucial for making impactful decisions related to resource allocation, service expansion, and policy adjustments.
+This information can be useful for making impactful decisions related to resource allocation, service expansion, and policy adjustments.
 Transit agencies can use this information to optimize fare structures, implement dynamic pricing models, or introduce discounts and incentives to attract more riders during specific times or for particular modes.
@@
 -   Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler
--   R Paper source code: https://github.com/thatonecalculator/math-28-project
+-   R Paper source code: <https://github.com/thatonecalculator/math-28-project>

### 2023-11-29 Paper.Rmd
Conclusion

--- 2023-11-27 Paper.Rmd
@@
 
+## Conclusion
+
+As previously stated, the aim of this study was to examine how the rapid growth of Los Angeles affects the demand for public transportation.
+Our quantitative research examines the elasticity of demand for transportation options in Los Angeles based on pricing, availability, convenience, and modes.
+With that, as this project was done with the principles of Business Calculus in mind, it is important to highlight and emphasize the importance of understanding the applications from our findings.
+
+Firstly, an improved understanding of the allocation of resources and labor is something that decision makers within the public transportation sector can take away from our analysis.
+Understanding the elasticity of demand helps city planners anticipate changes in transportation preferences and adjust infrastructure investments accordingly.
+On the projects page of the metro website (<https://www.metro.net/projects/>), there are dozens of various projects that each require valuable tax-dollars, time, resources, planning, and execution.
+As Los Angeles is the third largest metropolitan area by GDP, being efficient in its decision making surrounding the allocation of resources in all facets of public transportation is something that our analysis of the rates of change of the bus and rail ridership over a five-year period will help better inform.
+
+As anyone who has been to Los Angeles knows the horror stories of its infamous traffic, our analysis of the elasticity of demand for transportation options in Los Angeles can guide policies aimed at managing traffic congestion, an example of such being incentives for alternative transportation modes.
+As transportation is a significant contributor to air pollution and greenhouse gas emissions, understanding how demand for different transportation options impacts the environment can help in designing policies that encourage environmentally friendly modes of transportation, such as public transit or electric vehicles.
+In a more economic sense, the transportation sector is closely tied to the overall economy.
+Understanding the elasticity of demand can provide insights into the economic impact of specific changes in transportation options.
+This information is valuable for businesses, urban development, and job accessibility.
+
+\
+Overall, an analysis of the data of public transportation in a city like Los Angeles offers immense benefits.
+From the perspective of the city, having a clearer understanding of the elasticity and rates of change between transportation modes over a five-year period enables them to make better informed decisions compared to only looking at raw data and trusting government intuition.
+From a business and entrepreneurial perspective, understanding the historical and predicted demand of various transportation and pricing options allows for entrepreneurs to better understand the institutional voids that may exist in public transportation in Los Angeles, and to potentially develop the next problem solver.
+(reference proposed Union Station to Dodger Stadium 5,000 seat gondola estimated well over original \$300 million estimate ([latimes.com -Dodger Stadium Gondola](https://www.latimes.com/sports/story/2023-04-30/dodger-stadium-gondola-project-frank-mccourt))
+
+\newpage
+
 # Sources:

### 2023-11-29 Paper.Rmd
Updated Conclusion

@@
 output:
+  pdf_document:
+    extra_dependencies: amsmath
   html_document:
     df_print: paged
-  pdf_document:
-    extra_dependencies: amsmath
 abstract: This paper will examine how Los Angeles's demand for public transportation
@@
 As previously stated, the aim of this study was to examine how the rapid growth of Los Angeles affects the demand for public transportation.
-Our quantitative research examines the elasticity of demand for transportation options in Los Angeles based on pricing, availability, convenience, and modes.
+Our quantitative research examines the rate of change for transportation options in Los Angeles based on pricing, availability, convenience, and modes.
 With that, as this project was done with the principles of Business Calculus in mind, it is important to highlight and emphasize the importance of understanding the applications from our findings.
 
+\
+
 Firstly, an improved understanding of the allocation of resources and labor is something that decision makers within the public transportation sector can take away from our analysis.
-Understanding the elasticity of demand helps city planners anticipate changes in transportation preferences and adjust infrastructure investments accordingly.
+Understanding the rate of change helps city planners anticipate changes in transportation preferences and adjust infrastructure investments accordingly.
 On the projects page of the metro website (<https://www.metro.net/projects/>), there are dozens of various projects that each require valuable tax-dollars, time, resources, planning, and execution.
-As Los Angeles is the third largest metropolitan area by GDP, being efficient in its decision making surrounding the allocation of resources in all facets of public transportation is something that our analysis of the rates of change of the bus and rail ridership over a five-year period will help better inform.
+As Los Angeles is the third largest metropolitan area by GDP, being efficient in its decision making surrounding the allocation of resources in all facets of public transportation is something that our analysis of the rate of change of the bus and rail ridership over a five-year period will help better inform.
 
-As anyone who has been to Los Angeles knows the horror stories of its infamous traffic, our analysis of the elasticity of demand for transportation options in Los Angeles can guide policies aimed at managing traffic congestion, an example of such being incentives for alternative transportation modes.
+\
+
+As anyone who has been to Los Angeles knows the horror stories of its infamous traffic, our analysis of the rate of change for transportation options in Los Angeles can guide policies aimed at managing traffic congestion, an example of such being incentives for alternative transportation modes.
 As transportation is a significant contributor to air pollution and greenhouse gas emissions, understanding how demand for different transportation options impacts the environment can help in designing policies that encourage environmentally friendly modes of transportation, such as public transit or electric vehicles.
 In a more economic sense, the transportation sector is closely tied to the overall economy.
-Understanding the elasticity of demand can provide insights into the economic impact of specific changes in transportation options.
+Understanding the rate of change can provide insights into the economic impact of specific changes in transportation options.
 This information is valuable for businesses, urban development, and job accessibility.
@@
 Overall, an analysis of the data of public transportation in a city like Los Angeles offers immense benefits.
-From the perspective of the city, having a clearer understanding of the elasticity and rates of change between transportation modes over a five-year period enables them to make better informed decisions compared to only looking at raw data and trusting government intuition.
+From the perspective of the city, having a clearer understanding of the rate of change between transportation modes over a five-year period enables them to make better informed decisions compared to only looking at raw data and trusting government intuition.
 From a business and entrepreneurial perspective, understanding the historical and predicted demand of various transportation and pricing options allows for entrepreneurs to better understand the institutional voids that may exist in public transportation in Los Angeles, and to potentially develop the next problem solver.
@@
 -   Transit ridership data: Los Angeles Metro "L.A. METRO TRANSIT RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD", Patrick Chandler
+-   Dodger Stadium Gondola Project Proposal [latimes.com](https://www.latimes.com/sports/story/2023-04-30/dodger-stadium-gondola-project-frank-mccourt){.uri}
+-   Los Angeles Metro Projects [Metro Projects](https://www.metro.net/projects/){.uri}
 -   R Paper source code: <https://github.com/thatonecalculator/math-28-project>

### 2023-12-06 Paper.Rmd
Paragraph

@@
 As seen in the graph below, ridership data for both the Metro Bus system and Metro Light Rail between 2019 and 2023 has been heavily impacted by the 2020 pandemic.
-The bus ridership has had more of a sharp decline from 2019 to 2020 and bounce-back shortly, while the rail was slowing down from 2019 - 2021, with not a lot of recovery.
+Bus ridership had a sharp decline from 2019 to 2020, however, it bounced back shortly afterward.
+The graph below can give us a visual on the impact COVID-19 had on public transport in the city of Los Angeles.
+In April of 2020, bus ridership had decreased heavily to a low of below 5 million riders.
+By April 2021, it had recovered up to about 12.5 million riders.
+By the end of 2023, riders returned to just above 17 million.
+While the Metro Light Rail did not experience as sharp of a decline from 2019 - 2020, this decrease continued into 2021.
+That decline produced a low of below 5 million riders.
+In 2023, the number of riders has still not returned to pre-pandemic levis hovering at 5 million users.
+Both systems of transport saw reduction in users, although, the metro bus system saw a greater decline and greater recovery.
 

### 2023-12-06 Paper.tex
Paragraph

@@
 \author{Kainoa Kanter, Julian Jacobson, Aidan Allen}
-\date{November 21, 2023}
+\date{December 05, 2023}
 
@@
 derivatives to determine the rate of change in public transportation
-usage and studying the elasticity of demand for transportation options
-based on pricing, availability, convenience, and modes.
+usage.
 \end{abstract}
@@
 a quantitative assessment utilizing derivatives to determine the rate of
-change in public transportation usage. Additionally, the study examines
-the elasticity of demand for transportation options in Los Angeles based
-on pricing, availability, convenience, and modes. This analysis offers
-insights into the responsiveness of ridership to changes in these
-factors, informing strategies to enhance the public transportation
-system.
+change in public transportation usage. This analysis offers insights
+into the responsiveness of ridership to changes in these factors,
+informing strategies to enhance the public transportation system.
+
+This information can be useful for making impactful decisions related to
+resource allocation, service expansion, and policy adjustments. Transit
+agencies can use this information to optimize fare structures, implement
+dynamic pricing models, or introduce discounts and incentives to attract
+more riders during specific times or for particular modes. Additionally,
+knowing how pricing, availability, and convenience impact ridership
+allows for targeted marketing and communication strategies.
 
@@
 
-\hypertarget{elasticity-of-demand-analysis}{%
-\paragraph{Elasticity of Demand
-Analysis:}\label{elasticity-of-demand-analysis}}
-
-\begin{itemize}
-\item
-  Analyze the responsiveness of public transportation demand to changes
-  in price (fare changes), availability (service hours, frequency),
-  convenience (route changes, wait times), and modes (bus vs.~rail).
-\item
-  Employ econometric modeling to estimate the price elasticity of
-  demand, cross-elasticity, and income elasticity, using regression
-  analysis where ridership
-\end{itemize}
-
 \hypertarget{visualization}{%
@@
 
-\newpage
-
 \hypertarget{exploratory-data-analysis}{%
 \section{Exploratory Data Analysis}\label{exploratory-data-analysis}}
+
+As seen in the graph below, ridership data for both the Metro Bus system
+and Metro Light Rail between 2019 and 2023 has been heavily impacted by
+the 2020 pandemic. Bus ridership had a sharp decline from 2019 to 2020,
+however, it bounced back shortly afterward. The graph below can give us
+a visual on the impact COVID-19 had on public transport in the city of
+Los Angeles. In April of 2020, bus ridership had decreased heavily to a
+low of below 5 million riders. By April 2021, it had recovered up to
+about 12.5 million riders. By the end of 2023, riders returned to just
+above 17 million. While the Metro Light Rail did not experience as sharp
+of a decline from 2019 - 2020, this decrease continued into 2021. That
+decline produced a low of below 5 million riders. In 2023, the number of
+riders has still not returned to pre-pandemic levis hovering at 5
+million users. Both systems of transport saw reduction in users,
+although, the metro bus system saw a greater decline and greater
+recovery.
 
@@
 usage is increasing or decreasing over time.
+
+It is important to keep in mind the usage of information like such. For
+Metro, getting their ridership to pre-pandemic numbers would be a huge
+accomplishment in the public transit landscape of Los Angeles. Total
+ridership in May 2023 was at 77\% of 2019 levels, and in June 2023, it
+was at 81\% of its 2019 pre-pandemic level. When using March of 2020 as
+our reference point, we can analyze the potentially differing rates of
+change in the ridership of Metro Bus and Metro Rail usage over time.
+Over the first year from 2020, the total monthly ridership went down
+roughly by roughly 22.5 million users.
+
+\hfill\break
+
+The derivative of bus rider usage in Los Angeles from 2019 to 2022 shows
+a gradual recovery in ridership, with consistent year-over-year
+increases and efforts to restore pre-pandemic levels. The recovery of
+bus ridership in Los Angeles reflects the ongoing efforts of Metro to
+improve mobility and public transit accessibility in the region. These
+changes indicate a positive trend in bus rider usage, with a focus on
+restoring and surpassing pre-pandemic ridership levels.
 
@@
 
+\hypertarget{conclusion}{%
+\subsection{Conclusion}\label{conclusion}}
+
+As previously stated, the aim of this study was to examine how the rapid
+growth of Los Angeles affects the demand for public transportation. Our
+quantitative research examines the rate of change for transportation
+options in Los Angeles based on pricing, availability, convenience, and
+modes. With that, as this project was done with the principles of
+Business Calculus in mind, it is important to highlight and emphasize
+the importance of understanding the applications from our findings.~
+
+\hfill\break
+
+Firstly, an improved understanding of the allocation of resources and
+labor is something that decision makers within the public transportation
+sector can take away from our analysis. Understanding the rate of change
+helps city planners anticipate changes in transportation preferences and
+adjust infrastructure investments accordingly. On the projects page of
+the metro website (\url{https://www.metro.net/projects/}), there are
+dozens of various projects that each require valuable tax-dollars, time,
+resources, planning, and execution. As Los Angeles is the third largest
+metropolitan area by GDP, being efficient in its decision making
+surrounding the allocation of resources in all facets of public
+transportation is something that our analysis of the rate of change of
+the bus and rail ridership over a five-year period will help better
+inform.~
+
+\hfill\break
+
+As anyone who has been to Los Angeles knows the horror stories of its
+infamous traffic, our analysis of the rate of change for transportation
+options in Los Angeles can guide policies aimed at managing traffic
+congestion, an example of such being incentives for alternative
+transportation modes. As transportation is a significant contributor to
+air pollution and greenhouse gas emissions, understanding how demand for
+different transportation options impacts the environment can help in
+designing policies that encourage environmentally friendly modes of
+transportation, such as public transit or electric vehicles. In a more
+economic sense, the transportation sector is closely tied to the overall
+economy. Understanding the rate of change can provide insights into the
+economic impact of specific changes in transportation options. This
+information is valuable for businesses, urban development, and job
+accessibility.
+
+\hfill\break
+Overall, an analysis of the data of public transportation in a city like
+Los Angeles offers immense benefits. From the perspective of the city,
+having a clearer understanding of the rate of change between
+transportation modes over a five-year period enables them to make better
+informed decisions compared to only looking at raw data and trusting
+government intuition. From a business and entrepreneurial perspective,
+understanding the historical and predicted demand of various
+transportation and pricing options allows for entrepreneurs to better
+understand the institutional voids that may exist in public
+transportation in Los Angeles, and to potentially develop the next
+problem solver. (reference proposed Union Station to Dodger Stadium
+5,000 seat gondola estimated well over original \$300 million estimate
+(\href{https://www.latimes.com/sports/story/2023-04-30/dodger-stadium-gondola-project-frank-mccourt}{latimes.com
+-Dodger Stadium Gondola})
+
+\newpage
+
 \hypertarget{sources}{%
@@
   RIDERSHIP UP 10 PERCENT, SETS POST-PANDEMIC RECORD'', Patrick Chandler
+\item
+  Dodger Stadium Gondola Project Proposal
+  \href{https://www.latimes.com/sports/story/2023-04-30/dodger-stadium-gondola-project-frank-mccourt}{latimes.com}
+\item
+  Los Angeles Metro Projects
+  \href{https://www.metro.net/projects/}{Metro Projects}
+\item
+  R Paper source code:
+  \url{https://github.com/thatonecalculator/math-28-project}
 \end{itemize}
